Figures

### 2025-01-28 Figures/PaperFiguresextended.Rmd
Added comments

xxx double-check code
Supplementary Data Figure 1e
Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the proliferative samples with strict exclusion criteria with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
```{r}
dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")

p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_DEGsamples.rds"))

#Figure 1e
pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
)
dev.off()
```

@@

#Plot eigencorplots for the main annotations
plotting_PCA <- function(SUBSET) {
    p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_",SUBSET,".rds"))

    pdf(paste0(dir_out,"eigencorplot_final_BH_",SUBSET,".pdf"), width = 10, height = 5)
    eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
    )
    dev.off()
}

@@

Supplementary Figure 3a
Scores plot of the first two principal components PC1 and PC2 from sample-wise whole transcriptome PCA of the SCT assay, colored by endometriosis status.
```{r}
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# biplot
SFig3a <- biplot(p,
    colby = 'EndometriosisStatus',
    labSize = 0,drawConnectors = FALSE, 
    hline = 0, vline = 0,
    legendPosition = 'right')
ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
```

Supplementary Figure 3b and e
UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang
```{r}
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

#Menstrual Cycle Day
SFig3b <- plot_cells(epithelial_cells_monocle3, color_cells_by = "CycleDay",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
SFig3b
ggsave(filename = paste0(dir_out, 'SFig3b.pdf'),
       plot = SFig3b,
       width =20, height=9)

#Menstual Cycle Phase Wang
SFig3e <- plot_cells(epithelial_cells_monocle3 , color_cells_by = "WangMenstrualCyclePhase",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
SFig3e
ggsave(filename = paste0(dir_out, 'SFig3e.pdf'),
       plot = SFig3e,
       width =20, height=9)

@@

xxxsee above for object name in GEO
Supplementary Data Figure 3c
Look in IBU folder for object with trajectory line or A063/A066
```{r}
#Figure S3c epithelial cells
FeaturePlot
Fig3c <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCycleDay", raster = FALSE)
Fig3c
ggsave(filename = paste0(dir_out, 'Fig3c.pdf'),
       plot = Fig3c,
       width =20, height=9)
```

Supplementary Data Figure 3f
CODE OK
```{r}
#Figure 2b entire atlas
Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "Wang2020Phase", raster = FALSE)
Fig2b_1
ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
       plot = Fig2b_1,
       width =20, height=9)

@@

Supplementary Data Figure 3 g
CODE OK
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#cd endometriosis_endometrium_scRNA_atlas/_Data
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad

@@

Supplementary Data Figure 3 h
CODE OK
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#cd endometriosis_endometrium_scRNA_atlas/_Data
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad

@@

Supplementary Data Figure 4a
CODE OK
```{r plotting of cell types per sample}
##Barplots of cell type frequency split by the main 5 cell types

@@

#Make df with number of AnnotationRefined cells in each sample
df <- EndoAtlas_meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample, AnnotationMain, AnnotationRefined, EndometriosisGrade, MenstrualCyclePhase, epithelial.pseudotime, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup()

#Make bar plot
p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = sample(P63)) +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) +
  facet_grid(AnnotationMain ~ ., scales = "free_y")
p
#Save barplot
ggsave(paste0(dir_out, "SFig4a_barplot.pdf"),
       plot = p,
       width = 15, height = 6)

@@

Supplementary Data Figure 4b
CODE OK
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for samples without minor exclusion criteria

@@

#BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
ggplot(SFig4b, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7) +
  theme_bw() +
  expand_limits(y = 0)

ggsave(paste0(dir_out, "SFig4b_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 20, height = 12)

@@

Supplementary Data Figure 5a
```{r}
DEGs

@@

Supplementary Data Figure 5b-d
```{r}

>>>>

xxx double-check code
Supplementary Data Figure 1e
Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the proliferative samples with strict exclusion criteria with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
```{r}
# Set dir_out
dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
# Load CellChat object
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_DEGsamples.rds"))

# eigencorplot for Figure 1e
pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
)
dev.off()
```

@@

#Plot eigencorplots for the main annotations
plotting_PCA <- function(SUBSET) {
    # Load data    
    p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_",SUBSET,".rds"))
    # plot eigencorplot and save  
    pdf(paste0(dir_out,"eigencorplot_final_BH_",SUBSET,".pdf"), width = 10, height = 5)
    eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
    )
    dev.off()
}

@@

Supplementary Figure 3a
Scores plot of the first two principal components PC1 and PC2 from sample-wise whole transcriptome PCA of the SCT assay, colored by endometriosis status.
```{r}
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# biplot
SFig3a <- biplot(p,
    colby = 'EndometriosisStatus',
    labSize = 0,drawConnectors = FALSE, 
    hline = 0, vline = 0,
    legendPosition = 'right')
# save biplot
ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
```

Supplementary Figure 3b and e
UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang. UMAP of the entire dataset colored by the Symphony transferred Wang phases, show a clear visual separation between phase 1/2, phase 3, and phase 4. 
```{r}
# Load data
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

#Plot UMAP with Menstrual Cycle Day
SFig3b <- plot_cells(epithelial_cells_monocle3, color_cells_by = "CycleDay",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
SFig3b
# save UMAP
ggsave(filename = paste0(dir_out, 'SFig3b.pdf'),
       plot = SFig3b,
       width =20, height=9)

#Plot UMAP with Menstual Cycle Phase Wang
SFig3e <- plot_cells(epithelial_cells_monocle3 , color_cells_by = "WangMenstrualCyclePhase",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
SFig3e
# save UMAP
ggsave(filename = paste0(dir_out, 'SFig3e.pdf'),
       plot = SFig3e,
       width =20, height=9)

@@

Supplementary Data Figure 3c
Scatter plots showing the correlation of menstrual cycle day and serum progesterone with our menstrual cycle staging (c) and median Wang phases (d), transferred with Symphony (d).
```{r}
xxx missing

@@

Supplementary Data Figure 3f
UMAP of the entire dataset colored by the Symphony transferred Wang phases, show a clear visual separation between phase 1/2, phase 3, and phase 4. 
```{r}
## Figure 2b UMAP entire atlas
Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "WangMenstrualCyclePhase", raster = FALSE)
Fig2b_1
# Save UMAP
ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
       plot = Fig2b_1,
       width =20, height=9)

Supplementary Data Figure 3 g
Heatmap showing the expression of menstrual cycle phase markers (as in Fig. 2d) of the eutopic endometrial epithelial cells (g) from the Tan et al. 2022 single-cell atlas.
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#cd endometriosis_endometrium_scRNA_atlas/_Data
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad

@@

Supplementary Data Figure 3 h
Heatmaps showing the expression of menstrual cycle phase markers (as in Fig. 2d) of the eutopic endometrial mesenchymal cells (h) from the Tan et al. 2022 single-cell atlas.
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#cd endometriosis_endometrium_scRNA_atlas/_Data
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad

@@

Supplementary Data Figure 4a
Bar plot of cell type frequencies per sample, sorted by menstrual cycle phase and pseudo time and divided by the main 5 cell types. 
```{r plotting of cell types per sample}
##Barplots of cell type frequency split by the main 5 cell types

@@

#Make df with number of AnnotationRefined cells in each sample
SFig4a_df <- EndoAtlas_meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample, AnnotationMain, AnnotationRefined, EndometriosisGrade, MenstrualCyclePhase, epithelial.pseudotime, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup()

#Make bar plot
SFig4a <- ggplot(SFig4a_df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = sample(P63)) +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) +
  facet_grid(AnnotationMain ~ ., scales = "free_y")
SFig4a
#Save barplot
ggsave(paste0(dir_out, "SFig4a_barplot.pdf"),
       plot = SFig4a,
       width = 15, height = 6)

@@

Supplementary Data Figure 4b
Box plots of refined cell type frequencies between the main menstrual cycle phases. 
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for samples without minor exclusion criteria

@@

# BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
ggplot(SFig4b, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7) +
  theme_bw() +
  expand_limits(y = 0)
# Save BoxPlots
ggsave(paste0(dir_out, "SFig4b_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 20, height = 12)

@@

Supplementary Data Figure 5a
Genes underlying selected pathways enriched in the cluster- and stage-wise functional enrichment analysis.
```{r}
DEGs

@@

Supplementary Data Figure 5b-d
(b) Signaling pathways with relative information flow from CellChat analysis, ranked by overall information flow within the inferred networks between ENDO and non-ENDO.
(c) Circle plot of TGFb signaling pathway between all cell types in ENDO (top) and non-ENDO (bottom, no significant interactions).
(d) Bubble plot of selected communication probabilities with significant interaction probabilities (unadjusted p-value < 0.01)  within the TNF signaling pathway.
```{r}
#Load CellChat data
cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
